---
title: "Solutions"
date: "September 22nd, 2021"
output:
  html_document:
    theme: united
    highlight: tango
    toc: true
    toc_float: false
header-includes: \usepackage{mhchem}
---

## Solutions

Two types to discuss this semester:

* Aqeueous
* General solutions

```{r echo=FALSE, message=FALSE, warnings=FALSE}
properties <- c("solvent", "solute")
aqeueous_vals <- c("water", "Substance dissolved in the water (primarily ionic compounds/salts)")
general_vals <- c("Major component of the mixture", "Substance dissolved in the solvent")
df <- data.frame(properties, aqeueous_vals, general_vals)
colnames(df) <- c("Solutions", "Aqeueous", "General")
knitr::kable(df, escape = FALSE)
```

```{r echo=FALSE, message=FALSE, warnings=FALSE}
sol_types <- c("Strong electrolytes", "Non-electrolytes",  "Weak electrolytes")
rxns <- c("Acid/Base", "Precipitation",  "Reduction-Oxidation or Oxidation-Reduction (redox)")
eqns <- c("Molecular", "Total Ionic",  "?")
df <- data.frame(sol_types, rxns, eqns)
colnames(df) <- c("Solution Types", "Reactions", "Equations")
knitr::kable(df, escape = FALSE)
```

## Solution Types

**Strong electrolytic** (complete dissolution) solution

* Strong electrolytic solutions are good conductors of electricity
* We say that $NaCl$ is soluble in water: complete dissolution
  * Molecular Equation: $NaCl\text{ (s)} \rightarrow NaCl \text{ (aq)}$
* **Dissolution**: Ions that make up the compound *dissociate* from each other in solution.

$$
NaCl (s) \rightarrow Na^{+} (aq) + Cl^{-} (aq)
$$

**Non-electrolytic** solution

* Non-conductive of electricity
* Solutes of non-ionic compounds
* No dissociation (like dissolves like)

**Weak electrolytic** solution

* Partial dissociation of ions
* Acetic acid dissolves to an acetate ion and a hydrogen ion (the acid/proton)
* Poor conductors of electricity

## Acid Base Reactions

Arrhenius definition of acids and bases

* Acid is a source of $H^{+}\text{ (aq)}$ (proton), or hydronium $H_{3}O^{+}\text{ (aq)}$.
* Base is a source of $OH^{-}\text{ (aq)}$ (hydroxide anion).


### Strong Acids and Strong Bases

```{r echo=FALSE, message=FALSE, warnings=FALSE}
strong_acids <- c("$HCl$ hydrochloric", "$HBr$ Hydrobromic", "$HI$ Hydroiodic",
    "$HNO_{3}$ Nitric", "$HClO_{4}$ Perchloric", "$HClO_{3}$ Chloric acid", "$H_{2}SO_{4}$ Sulfuric", "")
strong_bases <- c("$LiOH$", "$NaOH$", "$KOH$", "$RbOH$", "$CsOH$", "$Ca(OH)_{2}$", "$Sr(OH)_{2}$", "$Ba(OH)_{2}$")
df <- data.frame(strong_acids, strong_bases, stringsAsFactors = FALSE)
colnames(df) <- c("Strong Acids", "Strong Bases")
knitr::kable(df, escape = FALSE)
```

* $HF$ is **not** a strong acid

## Redox Reaction

### Assigning Oxidation numbers

**Neutral elements and monatomic ions**:

- Neutral elements: the oxidation number equals zero
  - $Ne$, monatomic gas
  - $N_{2}$/$O_{2}$, diatomic molecules
  - $O_{3}$, triatomic
  - $Ag$, metal
- Monatomic ions: oxidation number equals charge on the ion
  - $Li^{+}$ has an oxidation number of 1
  - $Ba^{2+}$ has an oxidation number of 2
  - $N^{3-}$ has an oxidation number of -3

**Polyatomic species**:

- Neutral molecules: sum of oxidation numbers = 0
    - Carbon Monoxide ($CO$) has sum of O: $O$ is -2, $C$ is +2
    - Carbon Dioxide ($CO_{2}$) has sum of O: $O$ is -2, $C$ is +4 (there are two oxygens)
- Polyatomic ions: sum of oxidation numbers = charge of ion
    - Nitrite has a charge of -1 ($NO_{2}^{-}$). Oxygen is -2, Nitrogen is +3 (note: 2 oxygens, and must add up to -1)

### Priority in Assigning Oxidation Numbers

1. Group 1 Alkali Metals get $+1$
2. Group 2 Alkali Earth Metals get $+2$
3. Fluorine gets $-1$
4. Hydrogen gets $+1$
5. Oxygen gets $-2$
6. Group 17 (except F) gets $-1$
7. Group 16 (except O) gets $-2$
8. Group 15 gets $-3$

### Example: Assigning Oxidation Numbers

* $H_{2}$ (g) gets 0, it is a neutral diatomic molecule
* $MnO_{2}$ (s) as a sum gets 0, it is a neutral compound
  * $O$ is first, it gets -2
  * $Mn$ has to be +4
* $MnO_{4}^{-}$ (aq) as a sum gets -1, it is a polyatomic ion
  * $O$ goes first with -2
  * $Mn$ gets -7 (to sum to -1)
* $Fe$ (s) gets 0 (neutral monatomic)
* $Fe^{2+}$ (aq) gets +2
* $K_{2}Cr_{2}O_{7}$ (aq) gets 0 as a sum
  * $K$ gets +1
  * $O$ gets -2
  * $Cr$ gets +6
* $CClF_{3}$ gets 0 as a sum
  * $F$ gets -1
  * $Cl$ gets -1
  * $C$ gets +4
* $ClF_{3}$ gets 0 as a sum
  * $F$ gets -1
  * $Cl$ gets +3
* $H_{2}O$ gets 0 as a sum
  * $H$ gets +1
  * $O$ gets -2
* $H_{2}O_{2}$ (hydrogen peroxide) gets 0 as a sum
  * $H$ gets +1
  * $O$ gets -1
  * In peroxides, $O$ is -1.

### Redox reactions

Reduction-Oxidation and Oxidation-Reduction rxns are known as "redox" rxns.

* Redox rxns are where there ox. number (also called ox. state) of at least one element changes

Combusion Rxn:

$$
CH_{4} + 2O_{2} \rightarrow CO_{2} + 2H_{2}O
$$

```{r echo=FALSE, message=FALSE, warnings=FALSE}
elements = c("C", "H", "O")
reactant_side = c("-4", "+1", "0")
product_side = c("+4", "+1", "-2")
difference = c("+8", "0", "-2")
charge = c("8", "0", "-2")
df <- data.frame(elements, reactant_side, product_side, difference, charge, stringsAsFactors = FALSE)
colnames(df) <- c("Elements", "Reactant Side", "Product Side", "Difference", "Charge")
knitr::kable(df, escape = FALSE)
```

* **Reduction** is a _reduction_ in the oxidation number
  * In the example, this means Oxygen is reduced
* **Oxidation** is an _increase_ in the oxidation number
  * In the example, this means Carbon is oxidized

Associated idea:

* Oxidizing agent  
  * Thing that does the oxidation
  * Element that is reduced 
  * This is Oxygen in the example
* Reducing agent  
  * Thing that does the reduction
  * Element that is oxidized 
  * This is Carbon in the example

### Redox Example

Consider the rxn:

$$
2Al (s) + 3CuSO_{4} (aq) \rightarrow Al_{2}(SO_{4})_{3} (aq) + 3 Cu (s)
$$

Total ionic (aqeueous separate):

$$
2Al (s) + 3Cu^{2+} (aq) + 3SO_{4}^{2-} (aq) \rightarrow 2Al^{3+} + 3SO_{4}^{2-} (aq) + 3 Cu (s)
$$

Notice the $3SO_{4}^{2-}$ is a spectator ion, so we can write the _net ionic_ (excluding the spectator ion):

$$
2Al (s) + 3Cu^{2+} (aq)  \rightarrow 2Al^{3+} + 3 Cu (s)
$$


```{r echo=FALSE, message=FALSE, warnings=FALSE}
elems = c("Al", "Cu")
init_ox = c("0", "+2")
final_ox = c("+3", "0")
diff = c("+3", "-2")
change = c("Oxidation", "Reduction")
df <- data.frame(elems, init_ox, final_ox, diff, change, stringsAsFactors = FALSE)
colnames(df) <- c("Elements", "Initial Ox. Num", "Final Ox. Num", "Diff", "Change")
knitr::kable(df, escape = FALSE)
```

### Half Reaction

Half rxn is a balanced chemical equation that includes electrons to maintain charge balance.

Consider aluminum from the previous rxn:

$$
Al \rightarrow Al^{+3}
$$

Charge is imbalanced, must add electrons to the product side

$$
Al \rightarrow Al^{+3} + 3e^{-}
$$

Now, the charge is balanced.

Notice, that aluminum loses 3 electrons, so oxidation is loss of electrons 

Consider copper from the previous rxn:

$$
Cu^{2+} + 2e^{-} \rightarrow Cu
$$

Reactants need to gain 2 electrons, so reduction is gain of electrons

Acronym: OIL RIG (oxidation is loss of electrons) (reduction is gain of electrons)

$$
Al \rightarrow Al^{3+} + 3e^{-}
$$

$$
Cu^{2+} + 2e^{-} \rightarrow Cu
$$

Let's double the first and triple the second

$$
2Al \rightarrow 2Al^{3+} + 6e^{-}
$$

$$
3Cu^{2+} + 6e^{-} \rightarrow 3Cu
$$

Electrons cancel, so we are left with:

$$
2Al + 3Cu^{2+} \rightarrow 2Al^{3+} + 3Cu
$$

## Disproportionation Reaction

$$
2 H_{2}O_{2} \rightarrow + 2H_{2}O + O_{2}
$$

```{r echo=FALSE, message=FALSE, warnings=FALSE}
elements = c("H", "O")
init = c("+1", "-1")
final = c("+1", "-2/0")
change = c("0", "-1/+1")
df <- data.frame(elements, init, final, change, stringsAsFactors = FALSE)
colnames(df) <- c("Element", "Init", "Final", "Change")
knitr::kable(df, escape = FALSE)
```

A disproportionation rxn means **one element** is **both reduced and oxidized**.

## Concentration

We use **molarity** to figure out how much solute is in a solution

* $\text{Molarity} = M = \frac{\text{moles of solute}}{\text{Liters of solution (not solvent)}}$
* $\text{concentration} = \frac{n}{V}$ (same thing)
  * units are $\frac{\text{mol}}{\text{L}} = M$
  * A 1.0M solution means 1.0 moles of solute per liter

### Example: Concentration

1. What is the concentration of a solution that contains 0.715 moles of solute in 0.495L of solution?

$$
\text{conc} = \frac{n}{v} = \frac{0.715\text{ mol}}{0.495\text{ L}} = 1.44\text{ mol/L} = 1.44\text{ M}
$$

2. How many moles of $KI$ are in 85 mL of 0.50 M solution?

$$
\text{conc} = \frac{n}{v}
$$

$$
n = \text{conc} * V = 0.50\text{ M} * 0.085\text{ L} = 0.042\text{ mol KI}
$$

3. What volume of 1.23M $KClO_{3}$ solution contains 0.575 mol of solute?

$$
\frac{0.575}{1.23} = 0.467\text{ L}
$$

4. To add 135g of sucrose (molar mass 342.3 g/mol) to an experiment, how many mL of 3.30 M solution must be added?

a. How many of moles of sucrose does this mass represent?

$$
\frac{135}{} * \frac{1}{342.3} = 0.394\text{ mol}
$$

b. How many mL of 3.30M contain 0.394 mol?

$$
V = \frac{\text{mol}}{M} = \frac{0.394}{3.30} = 0.1195\text{ L} = 120. \text{ mL}
$$

### Stock Solution

Adding 127.4 g of $NaOH$ (40.00 g/mol) to make 500.0 mL of solution.

$$
\frac{127.4}{} * \frac{1}{40.00} = 3.185\text{ mol NaOH}
$$

$$
\frac{n}{v} = \frac{3.185}{.5} = 6.370\text{ M}
$$
